---
title: "Regression Analysis: MotorTrend_MPG"
author: "EHarris"
date: "7/19/2017"
output: html_document
---

```{r setup, include = FALSE, echo = TRUE, results = "hide"}
ipak <- function(pkg){
      new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
      if (length(new.pkg)) 
            install.packages(new.pkg, dependencies = TRUE)
      sapply(pkg, require, character.only = TRUE)
}

### Package Usage
packages <- c("plyr", "dplyr", "data.table", "dtplyr", "lubridate", "ggplot2", "scales",
                  "reshape2", "knitr", "R.cache", "stringr", "gtools", "quantreg",
                  "graphics", "corrplot", "broom")
ipak(packages)
```

### Overview  
Motor Trend has provided a data set, mtcars, with 32 observations from the 1974 Motor Trend magazine.  They wish to understand the relationship between a number of variables, predictors, and miles-per-gallon (MPG), outcome, Specifically, were are asked to answer the following two questions:

* Is an automatic or manual transmission better for MPG?
* Quantify the MPG difference between automatic and manual transmissions"  

Under a simple linear regression model, it would appear that a manual transmission adds more than 7 mpg over an automatic. Using additional variables to our regression model, we conclude that switching from an automatic to manual adds 1.8 mpg, all else equal. 

### Data Processing  
Given that this is an R dataset we can simply use data(mtcars) to load the dataset into the current enviromnent.  
```{r load_mtcars, include = TRUE, echo = TRUE}
data(mtcars)
mtcars$am <- as.factor(mtcars$am)
mtcars$cyl <- as.factor(mtcars$cyl) 
```


### Exploratory Data Analysis  
To obtain a high-level insight to the relationship of the different variables in the data set, we created a panel plot (see Appendix: Figure 1) of these relationships.  A correlation matrix (see Appendix: Figure 2) is also provided to provide a numeric value to the relationship.  We find that the variables cyl, disp, hp, and wt have the strongest relationship with mpg.  The variables drat, vs, and am are slightly less connected.  We use this information to help assess our regression analysis results.
```{r dataset_overview, include = FALSE, echo = FALSE, results = "hide"}
str(mtcars)
head(mtcars)
```


### Regression Analysis
The focus of our analysis is miles per gallon (MPG).  Starting with simple linear regression (mpg ~am), we add variables to build a more robust, and best fit, model.  We use several metrics to evaluate our model fit, including: adjusted r-square, residual squared error (sigma), and p-values.  We also use ANOVA and an analysis of the residuals to evaluate the model.

#### Regression Models
The initial model, 'fit1', looks at the effect of transmission type (automatic vs. manual) on the MPG of a car.  We also create a 'fit10' model that includes all 10 independent variables.  We perform a stepwise processs, adding & removing variables, to select the best variables.  This can be done manually and/or using a step() function which uses algorithm to evaluate whether variables contribute significantly to predicting mpg.

```{r regression_models, include = TRUE, echo = TRUE, results = "markup"}
fit1 <- lm(mpg ~ am, data = mtcars)
coef1 <- round(summary(fit1)$coef, 4)
glance1 <- round(glance(fit1)[1:6], 4)

fit10 <- lm(mpg ~ ., data = mtcars)
coef10 <- round(summary(fit10)$coef, 4)
glance10 <- round(glance(fit10)[1:6], 4)

bestfit <- step(fit10, direction = "both")
coef.best <- round(summary(bestfit)$coef, 4)
glance.best <- round(glance(bestfit)[1:6], 4)

## Obtained by manually adding / removing variables (using correlations as guide)
fit3 <- lm(mpg ~ am + wt + cyl, data = mtcars)
coef3 <- round(summary(fit3)$coef, 4)
glance3 <- round(glance(fit3)[1:6], 4)

fit4 <- lm(mpg ~ am + wt + cyl + hp, data = mtcars)
coef4 <- round(summary(fit4)$coef, 4)
glance4 <- round(glance(fit4)[1:6], 4)
```


#### Comparison of Regression Models
Two comparisons performed to evaluate the best model.  The first is a comparison of the r.square and p-value for each model.  Then, we perform ANOVA to evaluate whether model is significantly better.

##### Exhbit 1: R.Square and p-value Comparison
```{r regression_compare, include = TRUE, echo = TRUE, results = "markup"}
model <- c("fit1", "fit3", "fit4", "bestfit", "fit10")
model.compare <- rbind(glance1,glance3, glance4, glance.best, glance10)
model.compare <- cbind(model, model.compare)
model.compare
```

##### Exhibit 2: ANOVA
The findings provided below for Model 4: mpg ~ factor(am) + wt + factor(cyl) + hp produce a p-value of 0.026935, which indicates that this model is significant.  These variables further contribute to the accuracy of the model.

```{r model_compare_anova, include = TRUE, echo = TRUE, results = "markup"}
mpg.anova <- anova(fit1, fit3, fit4)
mpg.anova
```


### Residuals & Diagnostics
Below is a panel plot of residuals.  Understanding residuals is critical to understanding how well the regression model fits the data, goodness of fit.  The different plots provide some insight to how closely the regression line fits the data, explaining changes in mpg.

```{r plot_residuals, fig.keep = "high", fig.show = "asis", fig.path = 'figure/'}
par(mfrow = c(2,2), mar = c(5,4,1,1))
plot(fit4)
```


#### Residuals vs Fitted
This plot shows if residuals have non-linear patterns. The fact that the residuals are fairly well-distributed around the fitted line, and no particular pattern, we are fairly confident that the selected regression model is a relatively good fit. 


#### Normal Q-Q
This plot shows if residuals are normally distributed. It’s good fit if residuals are lined well on the straight dashed line.  In our plot, the residuals are fairly close to the straight line.  However, there appears to be a little greater distance for outliers, particularly upper right corner.  We may want to explore outliers a little further.


#### Scale-Location
We use the scale-location plot to check the assumption of equal variance (homoscedasticity).  Ideally, we would see points that are equally spread around a horizontal line along the range of predictors.  Although not a horizontal line, the fitted line does not show much slope up or down.  Additionally, the residual points are fairly evenly distributed about the line, no narrowing or spreading.

#### Residuals vs. Leverage
This is the final plot to examine.  The purpose of this plot is identify "influential" outliers.  If an outlier does not change, maybe minimally changes, the regression line whether it is included or excluded, the outlier is not "influential".  If outliers, residual points, appear in the upper right or lower right corner of the plot, thee points may be influential against the regression line.

### Additional Regression Diagnostics
In addition to plots provided above, we can look evaluate our model through some other tests.  Below, we look at a couple and explain the findings.

#### Hat Values




### Conclusion
Following our review of Exploratory Data Analysis, it was decided to test two assumptions.  From the boxplot, it appeared that the supplement 'OJ' promoted greater tooth growth than the supplement 'VC'.  It was also apparent that a higher dose promoted greater tooth growth, regardless of supplement introduced.  We chose to test these assumptions by setting NULL hypothesis that there was no difference in the tooth growth for supplement 'OJ' versus 'VC' and a higher dose versus lower dose.  The alternative hypothesis being that 'OJ' promotes greater tooth growth than 'VC' and a higher dose promotes greater tooth growth than a lower dose.

From our t.test, the p-values were rather small, with the exception of testing whether 'OJ' promotes greater tooth growth than 'VC' at a dose = 2.0, indicating that we reject the NULL hypothesis.  In summary, the analysis suggests that 'OJ' promotes greater tooth growth than 'VC'.  Additionally, regardless of the supplement selected, 'OJ' or 'VC', a higher dose of the supplement promotes greater tooth growth than lower dose.

### Appendix
The information provided below is an outline of the request or question driving this assingment and analysis of data for different cars and related miles per gallon (MPG).

#### Review criteria
1. Did the student interpret the coefficients correctly?
2. Did the student do some exploratory data analyses?
3. Did the student fit multiple models and detail their strategy for model selection?
4. Did the student answer the questions of interest or detail why the question(s) is (are) not answerable?
5. Did the student do a residual plot and some diagnostics?
6. Did the student quantify the uncertainty in their conclusions and/or perform an inference correctly?
7. Was the report brief (about 2 pages long) for the main body of the report and no longer than 5 with supporting appendix of figures?
8. Did the report include an executive summary?
9. Was the report done in Rmd (knitr)?


### Figure 1 (Panel Plot): Illustrate the relationship of each variable
This panel plot provides a quick, high-level, illustration of a variable within the dataset against each other variable in the dataset.  The pu
```{r panel_plot_exploratory, fig.keep = "high", fig.show = "asis", fig.path = 'figure/'}
pairs(mtcars, panel = panel.smooth, main = "MT Cars Data", col = 3)
```


#### Figure 2: Correlation Factor Matrix
```{r corr_matrix_grid, fig.keep = "high", fig.show = "asis", fig.path = 'figure/'}
par(mar = c(5,4,2,1))
mtcars.cor <- cor(mtcars)
corrplot(mtcars.cor, method = "number", type = "upper", add = FALSE, 
                  order = "original", is.corr = TRUE)
```


